### 2016-03-11 paper/icfp16-submission/semantics.tex
new sketch of lemma k-stuck

\begin{lem}
\label{lem:resolve-compat}
For all traces
$\trace \defeq \triple{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu},\ldots,\triple{e}{\vsu}{\tsu}$,
$\vcompat{\resolve{\ehole}{\vsu}}{\resolve{\thole}{\tsu}}$.
\end{lem}
\begin{proof}
  By induction on $\trace$.
  %
  In the base case $\trace = \triple{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu}$
  and $\vhole{\thole}$ is trivially compatible with $\thole$.
  %
  In the inductive case, consider the single-step extension of $\trace$,
  $\trace' = \trace,\triple{e'}{\vsu'}{\tsu'}$.
  %
  We show by case analysis on the evaluation rules that if
  $\vcompat{\resolve{\ehole}{\vsu}}{\resolve{\thole}{\tsu}}$, then
  $\vcompat{\resolve{\ehole}{\vsu'}}{\resolve{\thole}{\tsu'}}$.
  %
  \ES{fill in cases in the appendix}
\end{proof}

\begin{lem}
\label{lem:vsu-ext}
If  $\trace \defeq \triple{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu},\ldots$
and $\trace'$ is a single-step extension of $\trace$
and $\resolve{\vhole{\thole}}{\vsu} \neq \resolve{\vhole{\thole}}{\vsu'}$
then $\vsu' = \vsu[\ehole_1 \mapsto v_1] \ldots [\ehole_n \mapsto v_n]$.
\end{lem}
\begin{proof}
  By case analysis on the evaluation rules.
  %
  $\vhole{\thole}$ does not change, so if the resolved values differ then
  $\vsu \neq \vsu'$.
  %
  Note that only \forcesym can update \vsu\
  and only the first case,
  when we extend \vsu\ with $[\ehole \mapsto v]$,
  does.
  %
  \ES{fill in cases in the appendix}
\end{proof}

@@

\begin{lem}
\label{lem:k-stuck}
For all $t$,
if $\steptr{\trace}{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu}{e}{\vsu}{\tsu}$ and
   $\tincompat{t}{\ptype{\trace}{f}}$,
   then there exists a $v$ such that $\hastype{v}{t}$ and
   $\steps{\eapp{f}{v}}{\emptysu}{\emptysu}{\stuck}{\vsu'}{\tsu'}$ in at most
   $k$ steps, where $k$ is the length of $\trace$.
\end{lem}
\begin{proof}
  We can construct $v$ from $\trace$ as follows.
  %
  Let
  $$
  \trace_i = \triple{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu},
             \ldots,
             \triple{e_{i-1}}{\vsu_{i-1}}{\tsu_{i-1}},
             \triple{e_{i}}{\vsu_{i}}{\tsu_{i}}
  $$
  be the shortest prefix of $\trace$ such that
  $\tincompat{\ptype{\trace_i}{f}}{t}$.
  %
  We will show that $\resolve{\vhole{\thole}}{\vsu_{i-1}}$
  must contain some other hole $\ehole'$ which is instantiated with a
  concrete value at step $i$.
  %
  Furthermore, $\ehole'$ is instantiated in such a way that
  $\vincompat{\resolve{\vhole{\thole}}{\vsu_{i}}}{t}$.
  %
  Finally, we will show that if we had instantiated $\ehole'$ with a
  value such that
  $\vcompat{\resolve{\vhole{\thole}}{\vsu_{i}}}{t}$, the current
  step would have gotten $\stuck$.

  % By Lemma~\ref{lem:fixme} we know that
  % $\vcompat{\resolve{\vhole{\thole}}{\vsu}}{\resolve{\thole}{\tsu}}$.
  %
  By Lemma~\ref{lem:vsu-ext} we know that
  $\vsu_{i} = \vsu_{i-1}[\ehole_1 \mapsto v_1] \ldots [\ehole_n \mapsto v_n]$.
  %
  We will assume, without loss of generality, that
  $\vsu_{i} = \vsu_{i-1}[\ehole' \mapsto v']$.

  Next, we introduce a notion of \emph{value contexts}, defined
  analogously to evaluation contexts.
  %
  $$
    V ::= \bullet \spmid \vnode{t}{V}{v}{v} \spmid \vnode{t}{v}{V}{v} \spmid \vnode{t}{v}{v}{V}
  $$
  %
  Since $\vsu_{i-1}$ and $\vsu_{i}$ differ only in $\ehole'$ but the resolved
  values differ,
  $\resolve{\vhole{\thole}}{\vsu_{i-1}} = \invctx{\ehole'[\thole']}$
  and
  $\resolve{\vhole{\thole}}{\vsu_{i}} = \invctx{v'}$.
  %
  Furthermore, by Lemma~\ref{lem:resolve-compat} we know that
  $$
  \vcompat{\resolve{\vhole{\thole}}{\vsu_{i-1}}}{\resolve{\thole}{\tsu_{i-1}}}
  =
  \tcompat{\ptype{i-1}{f}}{t}
  $$
  $$
  \vcompat{\resolve{\vhole{\thole}}{\vsu_{i}}}{\resolve{\thole}{\tsu_{i}}}
  =
  \tincompat{\ptype{i}{f}}{t}
  $$
  %\ES{need to connect \ptype{\trace{i}}{f} and \resolve{\vhole{\thole}}{\vsu_i}}
  %
  Now, consider any value $u$ such that $\vcompat{\invctx{u}}{t}$,
  we show by case analysis on the evaluation relation that
  $\invctx{u}$ could not make progress at this step, and thus
  $\steps{\eapp{f}{\invctx{u}}}{\emptysu}{\emptysu}{\stuck}{\vsu'}{\tsu'}$.
\end{proof}

>>>>

\begin{lem}
\label{lem:resolve-compat}
For all traces
$\trace \defeq \triple{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu},\ldots,\triple{e}{\vsu}{\tsu}$,
$\vsub{\resolve{\thole}{\tsu}}{\resolve{\ehole}{\vsu}}$.
\end{lem}
\begin{proof}
  By induction on $\trace$.
  %
  In the base case $\trace = \triple{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu}$
  and $\thole$ is trivially a refinement of $\vhole{\thole}$.
  %
  In the inductive case, consider the single-step extension of $\trace$,
  $\trace' = \trace,\triple{e'}{\vsu'}{\tsu'}$.
  %
  We show by case analysis on the evaluation rules that if
  $\vsub{\resolve{\thole}{\tsu}}{\resolve{\ehole}{\vsu}}$, then
  $\vsub{\resolve{\thole}{\tsu'}}{\resolve{\ehole}{\vsu'}}$.
  %
  \ES{fill in cases in the appendix}
\end{proof}

\begin{lem}
\label{lem:vsu-ext}
If  $\trace \defeq \triple{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu},\ldots$
and $\trace'$ is a single-step extension of $\trace$
and $\resolve{\thole}{\tsu} \neq \resolve{\thole}{\tsu'}$
then $\tsu' = \tsu[\thole_1 \mapsto t_1] \ldots [\thole_n \mapsto t_n]$.
\end{lem}
\begin{proof}
  By case analysis on the evaluation rules.
  %
  $\thole$ does not change, so if the resolved types differ then
  $\tsu \neq \tsu'$.
  %
  Note that only \forcesym can change \tsu, and it can only do so
  via \unifysym, which can only extend \tsu.
  % and only the first case,
  % when we extend \vsu\ with $[\ehole \mapsto v]$,
  % does.
  %
  \ES{fill in cases in the appendix}
\end{proof}

@@

\begin{lem}
\label{lem:k-stuck}
For all $t$,
if $\steptr{\trace}{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu}{e}{\vsu}{\tsu}$ and
   $\tincompat{t}{\ptype{\trace}{f}}$,
   then there exists a $v$ such that $\hastype{v}{t}$ and
   $\steps{\eapp{f}{v}}{\emptysu}{\emptysu}{\stuck}{\vsu'}{\tsu'}$ in at most
   $k$ steps, where $k$ is the length of $\trace$.
\end{lem}
\begin{proof}
  We can construct $v$ from $\trace$ as follows.
  %
  Let
  $$
  \trace_i = \triple{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu},
             \ldots,
             \triple{e_{i-1}}{\vsu_{i-1}}{\tsu_{i-1}},
             \triple{e_{i}}{\vsu_{i}}{\tsu_{i}}
  $$
  be the shortest prefix of $\trace$ such that
  $\tincompat{\ptype{\trace_i}{f}}{t}$.
  %
  We will show that $\resolve{\thole}{\tsu_{i-1}}$
  must contain some other hole $\thole'$ that is
  instantiated at step $i$.
  %
  Furthermore, $\thole'$ is instantiated in such a way that
  $\tincompat{\resolve{\thole}{\tsu_{i}}}{t}$.
  %
  Finally, we will show that if we had instantiated $\thole'$ such that
  $\tcompat{\resolve{\thole}{\tsu_{i}}}{t}$, the current
  step would have gotten $\stuck$.

  % By Lemma~\ref{lem:fixme} we know that
  % $\vcompat{\resolve{\vhole{\thole}}{\vsu}}{\resolve{\thole}{\tsu}}$.
  %
  By Lemma~\ref{lem:vsu-ext} we know that
  $\tsu_{i} = \tsu_{i-1}[\thole_1 \mapsto t_1] \ldots [\thole_n \mapsto t_n]$.
  %
  We will assume, without loss of generality, that
  $\tsu_{i} = \tsu_{i-1}[\thole' \mapsto t']$.

  Next, we introduce a notion of type % and value
  contexts, defined
  analogously to evaluation contexts.
  %
  $$
  \begin{array}{lcl}
    T &::=& \bullet \spmid \ttree{T} \\
    % V &::=& \bullet \spmid \vnode{t}{V}{v}{v} \spmid \vnode{t}{v}{V}{v} \spmid \vnode{t}{v}{v}{V} \\
  \end{array}
  $$
  %
  Since $\tsu_{i-1}$ and $\tsu_{i}$ differ only in $\thole'$ but the resolved
  types differ,
  $\resolve{\thole}{\tsu_{i-1}} = \intctx{\thole'}$
  and
  $\resolve{\thole}{\tsu_{i}} = \intctx{t'}$.
  %
  Let $s$ be a concrete type such that $\intctx{s} = t$.
  %
  We show by case analysis on the evaluation rules that
  %
  $$\step{e_{i-1}}{\vsu_{i-1}}{\tsu_{i-1}[\thole' \mapsto s]}{\stuck}{\vsu}{\tsu}$$
  %
  Finally, by Lemma~\ref{lem:resolve-compat} we know that
  $\vsub{\resolve{\thole}{\tsu_{i-1}}}{\resolve{\ehole}{\vsu_{i-1}}}$,
  thus $\vsub{\intctx{\thole'}}{\resolve{\vhole{\thole}}{\vsu_{i-1}}}$.
  %
  Let $u = \gen{s}{\tsu}$
  and $v = \resolve{\vhole{\thole}}{\vsu_{i-1}}\sub{\ehole'[\thole']}{u}\sub{\thole'}{s}$,
  $\steps{\eapp{f}{v}}{\emptysu}{\emptysu}{\stuck}{\vsu}{\tsu}$ in $i$ steps.
  %and
  % $\vsub{\resolve{\thole}{\tsu_{i}}}{\resolve{\ehole}{\vsu_{i}}}$.
  % \vcompat{\resolve{\vhole{\thole}}{\vsu_{i-1}}}{\resolve{\thole}{\tsu_{i-1}}}
  % =
  % \tcompat{\ptype{i-1}{f}}{t}
  % \\
  % \vcompat{\resolve{\vhole{\thole}}{\vsu_{i}}}{\resolve{\thole}{\tsu_{i}}}
  % =
  % \tincompat{\ptype{i}{f}}{t}
  %
  % Finally, by the definition of $\typeofsym$,
  % we know that $\resolve{\vhole{\thole}}{\vsu_{i-1}} = \invctx{v}$
  % such that $\vsub{T'[\thole']}{v}$.
  %such that $\resolve{\typeof{v}}{\tsu_{i-1}} = T'[\thole']$
  %
  %\ES{need to connect \ptype{\trace{i}}{f} and \resolve{\vhole{\thole}}{\vsu_i}}
  %
  % Now, consider any value $u$ such that $\vcompat{\invctx{u}}{t}$,
  % we show by case analysis on the evaluation relation that
  % $\invctx{u}$ could not make progress at this step, and thus
  % $\steps{\eapp{f}{\invctx{u}}}{\emptysu}{\emptysu}{\stuck}{\vsu'}{\tsu'}$.
\end{proof}
